% Copyright (c) 2025 Intel Corporation.  All rights reserved.  See the file COPYRIGHT for more information.
% SPDX-License-Identifier: Apache-2.0

\documentclass[11pt]{article}

% Page layout
\usepackage[letterpaper, margin=1in]{geometry}
\setlength{\headheight}{14pt}
\usepackage{parskip}
\setlength{\parindent}{0pt}

% Typography
\usepackage{fancyvrb}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}

% Plotting
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% Code listings
\usepackage{listings}
\lstset{
    basicstyle=\ttfamily\small,
    columns=fullflexible,
    keepspaces=true,
    showstringspaces=false,
    breaklines=true,
    frame=none,
    xleftmargin=2em,
}

% Hyperlinks
\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue]{hyperref}

% Headers and footers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textit{Formal Verification for the Fulcrum CSP Toolchain}}
\fancyhead[R]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\title{\textbf{Formal Verification for the Fulcrum CSP Toolchain} \\[0.5em]
\large Feasibility Study, Project Plan, and Implementation Report}
\author{Mika Nystr\"om \and Claude (Anthropic)}
\date{DRAFT Version 0.7\\
\today}

\begin{document}

\maketitle

\vspace{1em}

\noindent
 Copyright \copyright\ 2025 Intel Corporation

 Licensed under the Apache License, Version 2.0 (the ``License'');
 you may not use this file except in compliance with the License.
 You may obtain a copy of the License at

 {\tt http://www.apache.org/licenses/LICENSE-2.0}

\tableofcontents
\newpage

%======================================================================
\section{Introduction}
%======================================================================

Asynchronous circuit design offers compelling advantages in power
efficiency, modularity, and electromagnetic compatibility, but it
introduces verification challenges that synchronous flows do not face.
The absence of a global clock means that correctness depends entirely on
the interplay of handshaking protocols: a single missing acknowledgment
or misordered communication can cause a physical deadlock that is
invisible to simulation until the right interleaving is exercised.

The Fulcrum CSP toolchain compiles Communicating Sequential Processes
(CSP) descriptions into cycle-accurate Modula-3 simulators.  Simulation
is effective for validating functional behavior, but it explores only a
subset of the reachable state space.  Formal verification---exhaustive
or symbolic exploration of \emph{all} reachable states---can provide
mathematical guarantees of deadlock freedom, livelock freedom, and
protocol conformance that simulation alone cannot.

This report assesses the feasibility of adding FDR4-style formal
verification capabilities to the Fulcrum CSP toolchain.  It surveys the
relevant theory, compares the Fulcrum CSP dialect with the CSPM input
language of FDR4, identifies favorable structural properties and
key challenges, proposes three implementation approaches, and
recommends an incremental path forward with a phased project plan.

%======================================================================
\section{Background}
%======================================================================

%----------------------------------------------------------------------
\subsection{CSP Theory and Semantic Models}
%----------------------------------------------------------------------

CSP (Communicating Sequential Processes), introduced by Hoare~\cite{hoare1985},
provides a process algebra for describing systems of concurrent agents
that interact by synchronous message passing over named channels.
The theory defines three progressively richer semantic models:

\begin{description}
\item[Traces model ($\mathcal{T}$).]
  A process is characterized by the set of finite sequences of
  events (traces) it can perform.  The traces model supports
  safety properties: ``nothing bad happens.''

\item[Stable failures model ($\mathcal{F}$).]
  A process is characterized by its traces together with its
  \emph{failures}---pairs $(s, X)$ where $s$ is a trace and $X$ is a
  set of events the process can refuse after performing~$s$.
  The failures model supports deadlock analysis and determinism
  checking.

\item[Failures-divergences model ($\mathcal{FD}$).]
  Extends the failures model with \emph{divergences}---traces after
  which the process can perform an infinite sequence of internal
  ($\tau$) actions.  This is the richest standard model and supports
  livelock analysis.
\end{description}

Refinement checking is the central verification technique: a system
$\mathit{Impl}$ satisfies a specification $\mathit{Spec}$ if
$\mathit{Impl}$ refines $\mathit{Spec}$ in the chosen model, written
$\mathit{Spec} \sqsubseteq \mathit{Impl}$.  In the traces model this
means the implementation produces no trace that the specification
forbids; in the failures model it additionally means the implementation
refuses no more than the specification allows.

%----------------------------------------------------------------------
\subsection{FDR4: Architecture and Algorithms}
%----------------------------------------------------------------------

FDR4~\cite{fdr4} (Failures-Divergences Refinement checker, version~4)
is the standard tool for CSP refinement checking.  Its architecture
comprises several stages:

\begin{enumerate}
\item \textbf{Parsing.}  The input language is CSPM (machine-readable
  CSP), a functional language with pattern matching, let-bindings,
  comprehensions, and built-in types (sets, sequences, tuples).
  Processes are defined using CSP operators: prefix ($\to$),
  external choice ($\Box$), internal choice ($\sqcap$), parallel
  composition ($\|$), interleaving ($|||$), hiding, and renaming.

\item \textbf{Supercombinator compilation.}  Process expressions are
  compiled into \emph{supercombinators}---compact representations that
  enumerate the transitions of a process as a function of its
  parameters.  This is the key innovation of FDR4: supercombinators
  enable on-the-fly state enumeration without materializing the full
  state graph up front.

\item \textbf{State exploration.}  FDR4 explores the product state
  space using BFS or DFS, driven by the supercombinator transition
  functions.  Compression techniques (diamond, sbisimulation,
  $\tau$-loop elimination) reduce the state space during exploration.

\item \textbf{Refinement checking.}  The tool checks
  $\mathit{Spec} \sqsubseteq_X \mathit{Impl}$ where $X$ is the
  chosen semantic model ($\mathcal{T}$, $\mathcal{F}$, or
  $\mathcal{FD}$).  Special-case checkers exist for deadlock freedom,
  livelock freedom, and determinism, which do not require an explicit
  specification process.
\end{enumerate}

FDR4 achieves scalability through several mechanisms: symbolic
representation of transitions, compression of the explored state
space, and multi-core parallel exploration.  Nevertheless, state
explosion remains the fundamental limit: a system of $n$ processes
each with $k$ states has a product space of up to $k^n$ states.

%======================================================================
\section{The Fulcrum CSP Toolchain}
%======================================================================

The Fulcrum CSP toolchain compiles hardware-oriented CSP descriptions
into native simulator binaries.  Its architecture is:

\begin{Verbatim}[samepage=true]
  .csp source          .sys system description
       |                        |
       v                        v
    cspfe (parser)       sys-file parser (Scheme)
       |                        |
       v                        v
  S-expression AST       system graph (channels,
  (.scm files)           process instances, ports)
       |                        |
       +-------+    +-----------+
               |    |
               v    v
     cspc (9-pass Scheme compiler)
               |
               v
     Modula-3 state machines
     (m3__*.m3, SimMain.m3, m3makefile)
               |
               v
     cm3 -build -override
               |
               v
     Native simulator binary (sim)
\end{Verbatim}

\subsection{Key Components}

\begin{description}
\item[\texttt{cspparse} (\texttt{cspfe}).]
  Lex/Yacc-based parser that reads \texttt{.csp} source files and
  emits an S-expression representation of the AST.  The AST is defined
  by the \texttt{cspgrammar} library.

\item[\texttt{cspgrammar}.]
  Modula-3 library defining the abstract syntax tree.  Key types
  include \texttt{CspExpression} (literals, binary/unary operators,
  channel operations like \texttt{Receive}, \texttt{Peek},
  \texttt{Probe}), \texttt{CspStatement} (assignment,
  \texttt{DetSelection}, \texttt{NondetSelection},
  \texttt{DetRepetition}, \texttt{Sequential}, \texttt{Parallel},
  \texttt{Send}, \texttt{Recv}), and \texttt{CspType} (integers,
  booleans, channels, arrays, structures).

\item[\texttt{cspc}.]
  The compiler driver.  Invoked with \texttt{-scm} it enters a Scheme
  REPL (via the embedded \texttt{mscheme} interpreter) that drives a
  9-pass compilation pipeline.  Scheme stubs expose the Modula-3 AST
  types to Scheme code, which performs type checking, optimization,
  and code generation.

\item[\texttt{cspsimlib}.]
  Modula-3 simulation runtime providing the scheduler, channel
  implementations (with configurable slack), process dispatch,
  and multi-threaded worker coordination.

\item[\texttt{cspbuild}.]
  A Scheme-level build system that reads \texttt{.sys} files
  (multi-process system descriptions), orchestrates compilation of
  each process, and generates the final simulator binary.
\end{description}

\subsection{Simulation Capabilities}

The generated simulator supports multiple scheduling strategies
(\texttt{-greedy}, \texttt{-nondet}, \texttt{-eager}), multi-threaded
execution (\texttt{-mt}~$N$), and distributed simulation
(\texttt{-master}/\texttt{-worker}).  It exercises one execution
path per run; randomized scheduling (\texttt{-nondet}) can improve
coverage but cannot guarantee exhaustive exploration.

%======================================================================
\section{Language Comparison: Fulcrum CSP vs.\ CSPM}
%======================================================================

The Fulcrum CSP dialect (sometimes called CAST) is designed for
hardware process descriptions, while CSPM is a general-purpose
process algebra notation.  The following table summarizes the key
differences:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Feature} & \textbf{Fulcrum CSP / CAST} & \textbf{CSPM (FDR4)} \\
\midrule
Channel model     & Bundled-data, directional & Abstract events \\
                  & (\texttt{L?x}, \texttt{R!x}) & (\texttt{c.v -> P}) \\
Data types        & Fixed-width integers, arrays, & Sets, sequences, tuples, \\
                  & structs & user-defined datatypes \\
Choice            & Deterministic (\texttt{[\,]}) & External ($\Box$), \\
                  & nondeterministic (\texttt{:[\,]:}) & internal ($\sqcap$) \\
Repetition        & \texttt{*[\,]} (infinite loop) & Recursion \\
Parallelism       & \texttt{||} (interleaved), & $\|$, $|||$, $[|\ldots|]$ \\
                  & implicit via \texttt{.sys} & with alphabet control \\
Hiding            & Not explicit & \texttt{P \textbackslash\ X} \\
Renaming          & Port bindings in \texttt{.sys} & \texttt{P[[a<-b]]} \\
Specification     & None (simulation only) & Refinement assertions \\
State space       & Implicit (simulated) & Explicit (enumerated) \\
\bottomrule
\end{tabular}
\end{center}

Despite the syntactic differences, the underlying computational model
is similar: both describe networks of sequential processes that
communicate via synchronous handshakes on named channels.  The Intel
dialect is closer to the ``blocking point'' formulation where a process
alternates between computation and a single blocking communication,
which is actually \emph{favorable} for verification---it constrains
the branching factor of the state graph.

%======================================================================
\section{Feasibility Analysis}
%======================================================================

%----------------------------------------------------------------------
\subsection{Favorable Factors}
%----------------------------------------------------------------------

Several properties of the Fulcrum CSP toolchain and the circuits it
describes make formal verification feasible:

\begin{description}
\item[Static topology.]
  The \texttt{.sys} file declares all processes and channels
  statically.  There is no dynamic process creation or channel
  allocation.  The system graph is known at compile time, which
  is a prerequisite for state enumeration.

\item[Bounded data.]
  Hardware processes operate on fixed-width integers (typically
  8--64~bits).  While the raw data domain is large, abstract
  interpretation or data-independence arguments can often reduce
  the effective data range.  Many protocol properties (deadlock
  freedom, channel protocol conformance) are independent of data
  values entirely.

\item[Blocking-point structure.]
  Fulcrum CSP processes alternate between local computation and a
  single blocking communication (send or receive on a channel).
  Between blocking points, the process executes deterministically.
  This means the LTS (Labelled Transition System) of each process
  has transitions only at communication points, dramatically reducing
  the number of states compared to a fine-grained interleaving
  semantics.

\item[Existing AST infrastructure.]
  The \texttt{cspgrammar} library provides a complete, well-typed
  AST.  The Scheme-accessible AST means that analysis passes can be
  prototyped rapidly in the existing \texttt{cspc} environment.

\item[Guarded-command normal form.]
  The \texttt{DetSelection} and \texttt{NondetSelection} statement
  types directly correspond to CSP external and internal choice.
  Guards are channel operations (\texttt{Probe}, \texttt{Receive}),
  which map naturally to LTS transitions.
\end{description}

%----------------------------------------------------------------------
\subsection{Challenges}
%----------------------------------------------------------------------

\begin{description}
\item[State explosion.]
  Even with the favorable blocking-point structure, realistic circuits
  have many processes.  The MiniMIPS reference design has 7~processes;
  a production pipeline might have 20--50.  With channel slack
  (buffering), each channel with slack~$d$ and width~$w$ adds up to
  $(2^w)^d$ states, though data abstraction can mitigate this.

\item[No specification language.]
  The Fulcrum CSP dialect has no mechanism for writing specifications or
  refinement assertions.  Deadlock freedom can be checked without a
  spec, but richer properties (e.g., ``the output channel produces
  values in sorted order'') require a specification language or
  assertion mechanism.

\item[Channel slack semantics.]
  FDR4 models channels as zero-buffer synchronous communications.
  The Intel toolchain supports non-unit channel slack (buffering),
  which must be modeled as explicit buffer processes in the LTS.
  This increases the state space but is straightforward to implement.

\item[Data-dependent control flow.]
  Some guards depend on data values (e.g., \texttt{x > 0}).
  Data-dependent branching requires either concrete enumeration of
  data values or symbolic techniques (SAT/SMT) to handle efficiently.

\item[Integration complexity.]
  FDR4 is a standalone tool written in Haskell with a C++ exploration
  engine.  Direct integration with the Modula-3/Scheme toolchain
  would require either a translation layer or a native implementation.
\end{description}

%======================================================================
\section{Proposed Approaches}
%======================================================================

We identify three approaches for adding formal verification to the
Fulcrum CSP toolchain, ranging from lowest implementation effort to
deepest integration.

%----------------------------------------------------------------------
\subsection{Approach~1: Translate to CSPM and Use FDR4}
%----------------------------------------------------------------------

Write a translator (in Scheme, within \texttt{cspc}) that converts
the compiled process network into CSPM syntax, emitting a
\texttt{.csp}\footnote{CSPM files conventionally use the \texttt{.csp}
extension; to avoid confusion with Intel \texttt{.csp} files, we would
use \texttt{.cspm} or \texttt{.fdr} as the extension.} file suitable
for FDR4.

\textbf{Advantages:}
\begin{itemize}
\item Leverages the mature, heavily optimized FDR4 engine.
\item Immediate access to all FDR4 analyses (deadlock, livelock,
  determinism, refinement).
\item Lowest implementation effort for basic functionality.
\end{itemize}

\textbf{Disadvantages:}
\begin{itemize}
\item FDR4 is proprietary software (University of Oxford); requires
  a license for commercial use.
\item Semantic gap: channel slack, data types, and probe operations
  require careful encoding.
\item Counterexample traces must be mapped back to the original
  Fulcrum CSP source, which adds complexity.
\item Maintenance burden: the translation must track changes in both
  the Fulcrum CSP dialect and CSPM.
\end{itemize}

%----------------------------------------------------------------------
\subsection{Approach~2: Native LTS Explorer}
%----------------------------------------------------------------------

Build a state-space exploration engine directly in the Fulcrum CSP
toolchain.  The engine would:

\begin{enumerate}
\item Compile each process to an LTS (states = blocking points,
  transitions = communication events).
\item Compose the per-process LTSs into a product LTS following the
  system topology declared in the \texttt{.sys} file.
\item Explore the product LTS using BFS/DFS with hash-based state
  storage, checking for deadlock (no enabled transitions) and
  optionally other properties.
\end{enumerate}

The natural implementation platform is Modula-3 (for performance)
with a Scheme API for scripting analyses, mirroring the existing
simulation architecture.

\textbf{Advantages:}
\begin{itemize}
\item No external tool dependencies; fully open-source.
\item Direct access to the AST and the simulation runtime's data
  structures (channels, process state).
\item Counterexample traces are in the native format.
\item Can exploit domain-specific optimizations: blocking-point
  reduction, data abstraction tailored to hardware channels.
\end{itemize}

\textbf{Disadvantages:}
\begin{itemize}
\item Significant implementation effort for the exploration engine.
\item Must implement compression techniques (partial-order reduction,
  symmetry reduction) to handle realistic circuits.
\item Initially limited to deadlock/livelock freedom; refinement
  checking requires substantially more infrastructure.
\end{itemize}

%----------------------------------------------------------------------
\subsection{Approach~3: SAT/BDD-Based Symbolic Model Checking}
%----------------------------------------------------------------------

Encode the process network as a Boolean transition relation and use
SAT solvers or BDD libraries for reachability analysis.  This is the
approach used by hardware model checkers (e.g., NuSMV, ABC).

\textbf{Advantages:}
\begin{itemize}
\item Handles large data widths efficiently (symbolic representation).
\item Mature tool ecosystem (MiniSat, CaDiCaL, CUDD).
\item Natural fit for hardware-oriented descriptions with fixed-width
  data.
\end{itemize}

\textbf{Disadvantages:}
\begin{itemize}
\item Largest implementation effort.
\item Encoding CSP concurrency semantics into Boolean relations is
  non-trivial and error-prone.
\item Loses the direct connection to CSP theory (traces, failures);
  properties must be expressed in CTL/LTL rather than as refinements.
\item Debugging counterexamples is harder (bit-level vs.\ process-level).
\end{itemize}

%======================================================================
\section{Recommended Approach}
%======================================================================

We recommend \textbf{Approach~2: a native LTS explorer}, starting
with deadlock freedom checking and incrementally adding capabilities.
The rationale is:

\begin{enumerate}
\item \textbf{No licensing constraints.}  FDR4's proprietary license
  is a significant obstacle for an open-source and Intel-internal
  toolchain.  A native implementation avoids this entirely.

\item \textbf{Domain-specific optimizations.}  The blocking-point
  structure of Fulcrum CSP processes means that each process's LTS is
  compact---typically tens to hundreds of states, not thousands.
  The main challenge is the product-space explosion, which can be
  addressed incrementally with partial-order reduction and
  compositional reasoning.

\item \textbf{Natural integration.}  The existing toolchain already
  compiles processes to state machines for simulation.  The same
  compilation pipeline can produce an LTS representation instead of
  (or in addition to) Modula-3 simulation code.  The Scheme scripting
  layer provides a natural interface for expressing verification
  queries.

\item \textbf{Incremental value.}  Deadlock freedom is the single
  most valuable property for asynchronous circuits, and it requires
  no specification language---only exhaustive exploration of the
  product state space.  This can be delivered as a first milestone
  with immediate practical value.

\item \textbf{Path to richer analyses.}  Once the LTS explorer
  exists, adding livelock detection (cycle detection on $\tau$-only
  paths) and trace refinement (product with a specification LTS) is
  a natural extension.
\end{enumerate}

The high-level architecture of the proposed verification engine is:

\begin{Verbatim}[samepage=true]
  .sys system description
           |
           v
    System graph (processes, channels, topology)
           |
           v
    Per-process LTS extraction
    (states = blocking points, transitions = chan ops)
           |
           v
    Product LTS composition (on-the-fly)
           |
           v
    +------+------+------+
    |      |      |      |
    v      v      v      v
  Deadlock  Livelock  Trace    Failures
  check     check     refine.  refine.
\end{Verbatim}

%======================================================================
\section{Project Plan}
%======================================================================

The project is organized into four phases, each delivering a usable
increment of verification capability.

%----------------------------------------------------------------------
\subsection{Phase~1: Per-Process LTS Extraction}
%----------------------------------------------------------------------

\textbf{Goal:} Extract an explicit LTS from a single compiled CSP
process.

\textbf{Deliverables:}
\begin{itemize}
\item A new compilation target in \texttt{cspc} that emits an LTS
  (set of states, initial state, set of labelled transitions) instead
  of Modula-3 simulation code.
\item States correspond to blocking points (channel send, receive,
  select).  Transitions are labelled with the channel operation
  (event).
\item Data variables are handled by bounded enumeration with
  configurable abstraction (e.g., collapse all values to a single
  abstract value for data-independent properties).
\item LTS output in a simple textual format (e.g., Aldebaran
  \texttt{.aut} or a custom S-expression format readable by
  \texttt{cspc}).
\item Scheme API: \texttt{(extract-lts \textit{process})} returns
  an LTS object.
\item Unit tests on the existing CSP test suite.
\end{itemize}

\textbf{Milestone:} Can extract and display the LTS of any single-process
CSP program (e.g., the producer and consumer from the \texttt{cspbuild}
examples).

%----------------------------------------------------------------------
\subsection{Phase~2: Product Composition and Deadlock Checking}
%----------------------------------------------------------------------

\textbf{Goal:} Compose per-process LTSs into a product system and
check for deadlock.

\textbf{Deliverables:}
\begin{itemize}
\item Product-state composition engine that synchronizes processes on
  shared channel events, following the topology from the \texttt{.sys}
  file.
\item Channel slack modeled as explicit FIFO buffer processes inserted
  between communicating processes.
\item On-the-fly BFS/DFS exploration with hash-based visited-state
  set.
\item Deadlock detection: a product state is a deadlock if no process
  can make progress (no enabled transitions in any component).
\item Counterexample generation: on deadlock, emit the trace
  (sequence of events) leading to the deadlocked state.
\item Scheme API: \texttt{(check-deadlock! \textit{sys-file})} returns
  either \texttt{\#t} (deadlock-free) or a counterexample trace.
\item Tested on the MiniMIPS reference design and the
  \texttt{cspbuild} examples.
\end{itemize}

\textbf{Milestone:} Can verify deadlock freedom of the producer-consumer
example and detect intentionally introduced deadlocks in test cases.

%----------------------------------------------------------------------
\subsection{Phase~3: State-Space Reduction}
%----------------------------------------------------------------------

\textbf{Goal:} Scale the explorer to handle realistic circuits
(20+ processes) through state-space reduction techniques.

\textbf{Deliverables:}
\begin{itemize}
\item \textbf{Partial-order reduction (POR):} Exploit the independence
  of events on disjoint channels to avoid exploring redundant
  interleavings.  The static topology from the \texttt{.sys} file
  directly provides the independence relation.
\item \textbf{Symmetry reduction:} Exploit structural symmetry in
  regular process arrays (e.g., pipeline stages, arbiters).
\item \textbf{Data abstraction:} Automatic abstraction of data
  channels for data-independent properties; user-guided abstraction
  for data-dependent properties.
\item \textbf{Compositional reasoning:} Verify subsystems independently
  and compose results, exploiting the hierarchical structure of
  asynchronous designs.
\item Performance benchmarks comparing exploration time and peak
  memory with and without each reduction.
\end{itemize}

\textbf{Milestone:} The MiniMIPS 7-process design is verified for
deadlock freedom within reasonable time and memory bounds (under
1~minute, under 1~GB).

%----------------------------------------------------------------------
\subsection{Phase~4: Livelock and Refinement Checking}
%----------------------------------------------------------------------

\textbf{Goal:} Extend the verification engine beyond deadlock freedom.

\textbf{Deliverables:}
\begin{itemize}
\item \textbf{Livelock detection:} Detect cycles in the product LTS
  that involve only internal ($\tau$) transitions and no visible
  progress.  Uses Tarjan's SCC algorithm on the $\tau$-reachable
  subgraph.
\item \textbf{Specification language:} A lightweight assertion
  mechanism, either embedded in \texttt{.sys} files or as a separate
  \texttt{.spec} file, for expressing trace properties (e.g.,
  ``channel~\texttt{out} always follows channel~\texttt{in}'').
\item \textbf{Trace refinement:} Check that the implementation's
  trace set is a subset of the specification's trace set, following
  the standard CSP refinement algorithm.
\item \textbf{Failures refinement} (stretch goal): Full
  failures-model refinement checking.
\item Documentation and user guide.
\end{itemize}

\textbf{Milestone:} Can express and verify a simple ordering property
on the producer-consumer example (``consumer receives values in the
order the producer sent them'').

%----------------------------------------------------------------------
\subsection{Timeline Summary}
%----------------------------------------------------------------------

\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Phase} & \textbf{Focus} & \textbf{Key Deliverable} & \textbf{Duration} \\
\midrule
1 & LTS extraction  & Per-process LTS from \texttt{cspc} & 6--8 weeks \\
2 & Deadlock checking & Product explorer + counterexamples & 8--10 weeks \\
3 & Reduction  & POR, symmetry, data abstraction & 8--12 weeks \\
4 & Livelock \& refinement & Spec language + refinement checker & 10--14 weeks \\
\bottomrule
\end{tabular}
\end{center}

Phases~1 and~2 together deliver the minimum viable product: a deadlock
checker for multi-process CSP systems.  Phases~3 and~4 extend
scalability and expressiveness incrementally.

%======================================================================
\section{Implementation Status}
%======================================================================

Phases~1 and~2 have been implemented in the \texttt{cspc} toolchain.
This section documents what was built, how it performs, and what
we learned.

%----------------------------------------------------------------------
\subsection{Phase~1: Per-Process LTS Extraction (Complete)}
%----------------------------------------------------------------------

LTS extraction is implemented as a Scheme pass within \texttt{cspc}.
The function \texttt{(extract-system-lts!~\textit{sys-file})} compiles
each process type referenced by a \texttt{.sys} file, runs the 9-pass
compiler to produce the intermediate representation, and then performs a
symbolic walk of the compiled state machine.  States correspond to
blocking points (channel sends and receives); transitions are labelled
with \texttt{(send~\textit{ch})} or \texttt{(recv~\textit{ch})} for
channel operations, and \texttt{tau} for internal (unobservable)
transitions such as guarded-repetition termination.

The per-process LTSs are compact.  For the dining philosophers
benchmark, the deterministic fork process has 5~states and 4~transitions;
the na\"ive philosopher has 5~states and 4~transitions.  This confirms
the blocking-point hypothesis: hardware CSP processes yield small LTSs.

%----------------------------------------------------------------------
\subsection{Phase~2a: Explicit-State Product Composition (Complete)}
%----------------------------------------------------------------------

The explicit-state deadlock checker composes per-process LTSs into a
product LTS on-the-fly using BFS.  Synchronization follows the
\texttt{.sys} topology: a \texttt{(send~\textit{ch})} transition in one
instance synchronizes with a \texttt{(recv~\textit{ch})} transition in
the connected instance; \texttt{tau} transitions are taken independently.
A product state is deadlocked if no component has an enabled transition.

When deadlock is found, the checker returns a counterexample trace
(sequence of synchronization events leading to the deadlocked state).
The BFS guarantees the trace is of minimum length.

The checker is implemented in Scheme (\texttt{product.scm}) and uses
the Modula-3 \texttt{RefSeq.T} type for the BFS queue.  It handles
all 15~acceptance tests, including $N = 2$ and $N = 3$ dining
philosophers with both deterministic and probe-based forks.

\textbf{Limitation:} The explicit-state approach enumerates product
states one at a time.  For $N = 8$ dining philosophers, the product
state space has $5^{16} \approx 1.5 \times 10^{11}$ states
(8~forks $\times$ 8~philosophers, each with 5~states), which is
infeasible for explicit enumeration.

%----------------------------------------------------------------------
\subsection{Phase~2b: BDD-Based Symbolic Deadlock Checking (Complete)}
%----------------------------------------------------------------------

To handle larger systems, a BDD-based symbolic deadlock checker was
implemented (\texttt{symbolic.scm}).  The checker encodes process states
as Boolean variables (2~bits per instance for a 5-state process)
and computes reachability via a frontier-based fixed-point iteration:
\[
  F_{k+1} = \mathit{Image}(F_k) \setminus R_k, \qquad
  R_{k+1} = R_k \cup F_{k+1}
\]

The transition relation uses \emph{partitioned transition relations}
rather than a single monolithic BDD.  Each partition covers either
one instance (tau transitions) or two instances (channel synchronization),
with no frame conditions for non-participating instances.  The image
is computed per-partition:
\[
  \mathit{Image}(F) = \bigcup_{i} \bigl(
    \exists V_i.\; F \wedge T_i(V_i, V'_i)
  \bigr)[V'_i / V_i]
\]
where $T_i$ mentions only the current- and next-state variables
$V_i, V'_i$ of the 1--2 participating instances.  Non-participating
instances' state passes through unchanged in the conjunction with $F$,
which implicitly provides the frame condition without building explicit
equivalence BDDs.  Each conjunction $F \wedge T_i$ produces a small
intermediate because $T_i$ involves only 4--8 BDD variables.

After convergence, deadlocked states are those in the reached set with
no successor: $\mathit{Deadlock} = \mathit{Reached} \wedge \neg \mathit{HasSucc}$,
where $\mathit{HasSucc} = \bigvee_i \exists V'_i.\; T_i(V_i, V'_i)$.
Operation caches are cleared between iterations via
\texttt{BDD.ClearCaches} to reclaim memory from small intermediates.

The implementation uses the existing \texttt{bdd} library (part of
the CM3 tree).  All BDD operations are accessible from Scheme via
auto-generated stubs (\texttt{SchemeStubs}).

\textbf{Results.}
The symbolic checker handles $N = 8$ dining philosophers (16~instances,
32~BDD variables, 32~partitions of 16--18~nodes each) and reaches the
fixed point in 8~iterations.  The following table shows the BDD size
of the reached set at each iteration:

\begin{center}
\begin{tabular}{rrr}
\toprule
\textbf{Iteration} & \textbf{Reached BDD} & \textbf{Frontier BDD} \\
\midrule
0 & 194 & 192 \\
1 & 474 & 452 \\
2 & 866 & 756 \\
3 & 1,230 & 896 \\
4 & 1,454 & 756 \\
5 & 1,542 & 452 \\
6 & 1,562 & 192 \\
7 & 1,564 & 66 \\
8 (fixed point) & 1,564 & 1 \\
\bottomrule
\end{tabular}
\end{center}

The BDD representation is compact: 1,564~nodes encode the full
reachable state set (which contains an exponential number of concrete
states).  The deadlock BDD has 66~nodes.  Peak memory is
approximately 217\,MB on Apple~M2.

%----------------------------------------------------------------------
\subsection{Memory Optimization}
%----------------------------------------------------------------------

The initial monolithic implementation (a single BDD for the full
transition relation~$T$) used approximately 2.7\,GB peak memory for
$N = 8$ dining philosophers, because the conjunction
$F \wedge T$ exploded into a large intermediate BDD.

\textbf{BDD garbage collection.}
Four procedures were added to the \texttt{BDD} interface for explicit
memory management: \texttt{ClearCaches} (drops operation caches),
\texttt{MarkRoot}/\texttt{UnmarkRoots} (GC root set management),
and \texttt{CollectGarbage} (mark-and-sweep).  However, GC between
iterations had no effect on peak memory (2,749\,MB vs.\ 2,702\,MB)
because the peak occurs \emph{within} an iteration during the
monolithic conjunction.

\textbf{Partitioned transition relations.}
The solution was to never form the monolithic~$T$.  Each transition
is kept as a separate partition BDD involving only the 1--2
participating process instances (16--18 BDD nodes per partition).
Frame conditions for non-participating instances are omitted entirely:
by conjoining the frontier directly with a partition and quantifying
out only the participating current-state variables, the other
instances' state passes through unchanged.

\begin{center}
\begin{tabular}{lrr}
\toprule
& \textbf{Monolithic $T$} & \textbf{Partitioned} \\
\midrule
Transition BDD    & 2,307 nodes (1 BDD)   & 16--18 nodes $\times$ 32 partitions \\
Peak RSS          & 2,702\,MB             & 217\,MB \\
Memory reduction  & ---                   & \textbf{12.4$\times$} \\
Result            & DEADLOCK              & DEADLOCK \\
\bottomrule
\end{tabular}
\end{center}

The partitioned approach also calls \texttt{BDD.ClearCaches} between
iterations to reclaim operation cache memory from the many small
intermediates produced by the per-partition image computation.

%----------------------------------------------------------------------
\subsection{Phase~2c: MDD/Saturation-Based Symbolic Deadlock Checking}
%----------------------------------------------------------------------

While the BDD-based checker reduces memory by 12$\times$ over the
monolithic approach, it still uses 217\,MB for $N = 8$ dining
philosophers and would require gigabytes for $N = 16$.  The fundamental
limitation is that BDDs encode each process instance using
$\lceil \log_2 k \rceil$ Boolean variables (where $k$ is the number of
local states), and the image computation must conjoin the frontier with
transition relations across \emph{all} Boolean variables at once.

Multi-valued Decision Diagrams (MDDs)~\cite{ciardo2001} offer a
fundamentally different approach.  An MDD has one level per process
instance, where each internal node at level~$i$ has $k_i$ children
(one per local state of instance~$i$).  This eliminates the binary
encoding overhead and enables the \emph{saturation algorithm}
(Ciardo, Marmorstein, and Siminiceanu~\cite{ciardo2001}), which fires
transition events bottom-up, level by level, achieving local fixed
points before composing upward.

\subsubsection{MDD Library}

A new \texttt{csp/mdd/} package implements the MDD data structure and
saturation algorithm:

\begin{description}
\item[\texttt{MDD.i3}, \texttt{MDDImpl.m3}.]
  Quasi-reduced MDD forest with hash-consing (canonical nodes via
  per-level unique tables).  Operations: \texttt{Union},
  \texttt{Intersection}, \texttt{Difference}, \texttt{MakeNode},
  \texttt{Singleton}.  Direct-mapped operation caches for set
  operations.

\item[\texttt{MDDEvent.i3/.m3}.]
  Kronecker-encoded events.  Each transition event is represented
  by sparse \texttt{(from, to)} matrices at the 1--2 affected MDD
  levels (identity at all intervening levels).  Tau events affect
  one level; synchronization events affect two levels (the sender
  and receiver instances).

\item[\texttt{MDDSaturation.i3/.m3}.]
  The saturation algorithm.  Events are partitioned by their top level
  and fired bottom-up.  At each level, all events whose top level
  matches are fired repeatedly until a fixed point is reached.
  A direct-mapped saturation cache (65,536 entries) records already-saturated
  nodes.  The \texttt{HasSuccessor} procedure computes the subset of
  states with at least one enabled transition, enabling deadlock
  detection as $\mathit{Deadlock} = \mathit{Reached} \setminus
  \mathit{HasSucc}$.
\end{description}

The library is integrated into \texttt{cspc} via auto-generated Scheme
stubs (\texttt{SchemeStubs}) for all three modules.  The Scheme entry
point is \texttt{(check-deadlock-saturation! \textit{sys-file})}, implemented
in \texttt{saturation.scm}.

\subsubsection{Benchmark Results}

\begin{center}
\begin{tabular}{llrrrr}
\toprule
\textbf{System} & \textbf{Method} & \textbf{Nodes} & \textbf{Peak RSS} & \textbf{Time} & \textbf{Result} \\
\midrule
Dining $N{=}2$ det & Saturation & 9 & --- & $< 1$\,s & DEADLOCK \\
Dining $N{=}3$ det & Saturation & 21 & --- & $< 1$\,s & DEADLOCK \\
Dining $N{=}3$ asym & Saturation & 26 & --- & $< 1$\,s & deadlock-free \\
\midrule
Dining $N{=}8$ det & BDD (partitioned) & 1,564 & 217\,MB & 3.7\,s & DEADLOCK \\
Dining $N{=}8$ det & MDD saturation & 765 & \textbf{38\,MB} & \textbf{2.1\,s} & DEADLOCK \\
\midrule
Dining $N{=}16$ det & BDD & \multicolumn{3}{c}{\textit{impractical}} & --- \\
Dining $N{=}16$ det & MDD saturation & 196,605 & 233\,MB & 7.2\,s & DEADLOCK \\
\bottomrule
\end{tabular}
\end{center}

For $N = 8$, the MDD/saturation approach uses \textbf{5.7$\times$ less
memory} (38\,MB vs.\ 217\,MB) and runs \textbf{1.8$\times$ faster}
(2.1\,s vs.\ 3.7\,s) than the BDD approach with partitioned transition
relations.  More importantly, the MDD approach scales to $N = 16$
(32~instances, 64~events, 32~MDD levels) in 7.2~seconds and 233\,MB,
a problem size that is impractical for BDDs.

The MDD representation is also more compact: 765~MDD nodes for $N = 8$
vs.\ 1,564~BDD nodes.  This is because each MDD level directly
represents a process instance's state as a multi-valued child array,
whereas the BDD must decompose each state into binary variables.

The 50 unit tests in the MDD test suite (\texttt{csp/mdd/test/})
cover terminal nodes, singletons, set operations, quasi-reduced form,
hash-consing, event construction, and saturation on mutex,
producer-consumer, and dining philosopher systems.

\subsubsection{Saturation Algorithm Correctness}

The key algorithmic insight is the mutual recursion between
\texttt{Saturate}, \texttt{SatFire}, and \texttt{RecFire}.  At each
MDD level, events are fired until a fixed point.  When an event fires
and produces new states, the merged node is re-saturated at that level.
Termination is guaranteed because the state set can only grow (Union
is monotone) and the total number of reachable states is finite.

A subtle correctness issue arose during implementation: when
\texttt{SatFire} at the top level of a sync event produces fired
states that are already present in the MDD, the result of
\texttt{Union(node, fired)} is the same node (by hash-consing).
Without checking for this, \texttt{SaturateLocal} would be called
recursively with an identical node, causing infinite mutual recursion
(manifesting as a stack overflow on ARM64 macOS).  The fix is a
single identity check: \texttt{IF merged = node THEN RETURN node END}.

%----------------------------------------------------------------------
\subsection{Summary of Verification Approaches}
%----------------------------------------------------------------------

\begin{center}
\begin{tabular}{lrrrl}
\toprule
\textbf{Method} & \textbf{$N{=}2$} & \textbf{$N{=}8$} & \textbf{$N{=}16$} & \textbf{Scalability} \\
\midrule
Explicit-state (BFS) & $< 1$\,s & \textit{infeasible} & \textit{infeasible} & $O(k^n)$ states \\
BDD (partitioned) & $< 1$\,s & 3.7\,s / 217\,MB & \textit{impractical} & exponential BDD growth \\
MDD saturation & $< 1$\,s & 2.1\,s / 38\,MB & 7.2\,s / 233\,MB & best for async systems \\
\bottomrule
\end{tabular}
\end{center}

The MDD/saturation approach is the clear winner for asynchronous CSP
systems.  Its level-by-level fixed-point computation exploits the
locality of asynchronous events (each event affects only 1--2 process
instances), which is exactly the structure of CSP systems composed via
channel synchronization.

%----------------------------------------------------------------------
\subsection{Phase~3a: Partial-Order Reduction (Complete)}
%----------------------------------------------------------------------

The explicit-state deadlock checker explores all interleavings of
enabled transitions, which is exponential in the number of processes.
Partial-order reduction (POR)~\cite{por} exploits the observation that
many interleavings are equivalent for deadlock checking: if two
transitions touch disjoint sets of process instances, their execution
order does not affect whether a deadlock is reached.

\subsubsection{Independence Relation from Channel Topology}

In the CSP channel model, the independence relation follows directly
from the \texttt{.sys} topology:
\begin{itemize}
\item A \emph{tau} transition on instance~$i$ has instance-set $\{i\}$.
\item A \emph{sync} transition on channel~$c$ between sender~$s_i$ and
  receiver~$r_i$ has instance-set $\{s_i, r_i\}$.
\item Two transitions are \emph{independent} iff their instance-sets
  are disjoint.
\end{itemize}

No runtime dependency analysis is needed---the static channel topology
provides the independence relation at system construction time.

\subsubsection{Ample Set Selection}

The POR implementation (\texttt{por.scm}) uses the \emph{stubborn set}
method.  At each product state, instead of exploring all enabled
transitions, the checker selects an \emph{ample set}---a subset
sufficient for deadlock detection.  The selection strategy:

\begin{enumerate}
\item For each process instance~$i$, collect all transitions touching~$i$.
\item Check whether all transitions of~$i$ are independent of all
  transitions \emph{not} touching~$i$.
\item If so, the transitions of~$i$ form a valid ample set.
\item Pick the instance giving the smallest ample set.
\item If no reduction is possible, fall back to the full transition set.
\end{enumerate}

For deadlock detection, only conditions C0 (non-emptiness) and C1
(independence closure) are required.  The cycle condition~C3 (needed
for liveness properties) is not needed because deadlock detection
examines only fully expanded states with no successors.

\subsubsection{Implementation}

The function \texttt{check-deadlock-por!} in \texttt{por.scm}
follows the same BFS structure as \texttt{check-deadlock!} but calls
\texttt{product-successors-por} (which annotates each transition with
its instance-set) followed by \texttt{select-ample-set} to reduce the
transitions explored at each state.  Counterexample tracing works
unchanged---the BFS parent map is still valid with reduced exploration.

Statistics are reported at completion: total transitions generated
vs.\ transitions after POR selection, with the percentage reduction.

\subsubsection{Benchmark Results}

\begin{center}
\begin{tabular}{llrrrl}
\toprule
\textbf{System} & \textbf{Method} & \textbf{States} & \textbf{Trans.} & \textbf{Reduction} & \textbf{Result} \\
\midrule
Producer-consumer & Full BFS     & 4  & 4  & ---  & deadlock-free \\
                  & POR          & 3  & 3  & 25\% & deadlock-free \\
\midrule
Dining $N{=}2$ det & Full BFS   & 4  & 3  & ---  & DEADLOCK \\
                    & POR        & 3  & 2  & 33\% & DEADLOCK \\
\midrule
Dining $N{=}3$ det & Full BFS   & 8  & 6  & ---  & DEADLOCK \\
                    & POR        & 4  & 3  & 50\% & DEADLOCK \\
\midrule
Dining $N{=}3$ asym & Full BFS  & 33 & 21 & ---  & deadlock-free \\
                     & POR       & 14 & 14 & 33\% & deadlock-free \\
\midrule
Dining $N{=}8$ det  & Full BFS  & \multicolumn{3}{c}{\textit{infeasible}} & --- \\
                     & POR       & 9  & 8  & 77\% & DEADLOCK \\
\bottomrule
\end{tabular}
\end{center}

POR is most effective when processes have independent transitions---the
dining $N{=}8$ case achieves 77\% transition reduction because each
philosopher's initial fork-acquire is independent of non-adjacent
philosophers.  Notably, POR makes the explicit-state checker feasible
for $N{=}8$ dining philosophers (9~states explored vs.\ infeasible
without POR), since the deadlock is reached along a short path where
each philosopher acquires one fork in sequence.

%----------------------------------------------------------------------
\subsection{Phase~3b: MDD Level Ordering (Complete)}
%----------------------------------------------------------------------

The saturation algorithm's performance depends critically on the MDD
level ordering.  A synchronization event between instances at MDD
levels~$\ell_1$ and~$\ell_2$ has \emph{span} $|\ell_1 - \ell_2|$.
Events with large span require the saturation algorithm to traverse
many intermediate identity levels, increasing both time and MDD node
count.

\subsubsection{Greedy BFS Ordering}

The implementation uses a FORCE-like greedy heuristic based on BFS
traversal of the communication graph:
\begin{enumerate}
\item Start with the instance having the most connections (center of
  mass).
\item BFS outward, placing each newly visited neighbor at the next
  available level.
\item This tends to minimize the sum of channel spans across all sync
  events.
\end{enumerate}

\subsubsection{Integration}

The ordering is computed by \texttt{compute-level-ordering-from-channel-map}
in \texttt{saturation.scm}, which is called after the channel map is
built.  The function returns a permutation mapping each instance index
to its MDD level.  This replaces the previous identity ordering.

\subsubsection{Benchmark Results}

\begin{center}
\begin{tabular}{llrrrr}
\toprule
\textbf{System} & \textbf{Ordering} & \textbf{MDD nodes} & \textbf{Peak RSS} & \textbf{Time} & \textbf{Improvement} \\
\midrule
Dining $N{=}8$ det  & Identity (prev.)  & 765     & 38\,MB  & 2.1\,s & --- \\
                     & Greedy BFS        & 36      & 38\,MB  & 2.0\,s & \textbf{21$\times$ fewer nodes} \\
\midrule
Dining $N{=}16$ det & Identity (prev.)  & 196,605 & 233\,MB & 7.2\,s & --- \\
                     & Greedy BFS        & 72      & 47\,MB  & 2.7\,s & \textbf{2,731$\times$ fewer nodes} \\
\bottomrule
\end{tabular}
\end{center}

The greedy BFS ordering produces dramatic improvements.  For $N = 8$,
the MDD shrinks from 765 to 36~nodes (21$\times$).  For $N = 16$,
the improvement is 2,731$\times$: from 196,605 to just 72~MDD nodes,
with memory dropping from 233\,MB to 47\,MB and time from 7.2\,s to
2.7\,s.  The key insight is that the BFS walk from a central instance
naturally places communicating partners at adjacent MDD levels, keeping
synchronization event spans small (1--2 levels rather than up to $n$).

%----------------------------------------------------------------------
\subsection{Updated Summary of Verification Approaches}
%----------------------------------------------------------------------

\begin{center}
\begin{tabular}{lrrrrl}
\toprule
\textbf{Method} & \textbf{$N{=}2$} & \textbf{$N{=}8$} & \textbf{$N{=}16$} & \textbf{$N{=}1024$} & \textbf{Scalability} \\
\midrule
Explicit BFS         & $< 1$\,s & \multicolumn{3}{c}{\textit{infeasible}} & $O(k^n)$ states \\
Explicit BFS + POR   & $< 1$\,s & $< 1$\,s & \multicolumn{2}{c}{\textit{infeasible}} & reduced interleavings \\
BDD (partitioned)    & $< 1$\,s & 3.7\,s   & \multicolumn{2}{c}{\textit{impractical}} & exponential BDD growth \\
MDD (identity ord.)  & $< 1$\,s & 2.1\,s   & 7.2\,s & \textit{N/A} & poor level ordering \\
MDD + level + memo   & 1.6\,s   & 1.3\,s   & 1.5\,s & 7.2\,min & \textbf{best overall} \\
\bottomrule
\end{tabular}
\end{center}

The combination of MDD saturation with greedy BFS level ordering is
the clear winner.  Level ordering reduces $N = 16$ from 7.2\,s / 233\,MB
to 2.7\,s / 47\,MB---a 5$\times$ memory reduction and 2.7$\times$
speedup---while POR makes the explicit-state checker viable for
moderately large systems by eliminating redundant interleavings.

%----------------------------------------------------------------------
\subsection{Phase~3c: Deadlock Detection Optimization}
%----------------------------------------------------------------------

The initial scaling study revealed that the \texttt{HasSuccessor}
computation was the bottleneck, timing out at $N = 32$ even though
saturation itself completed in milliseconds.  Two optimizations
eliminated this bottleneck:

\begin{enumerate}
\item \textbf{Memoization.}  The \texttt{HasSuccForEvent} function
  recursively traverses the MDD, but shared sub-nodes caused the same
  computation to be repeated exponentially many times.  Adding a
  per-event computation cache (direct-mapped, 64K entries, keyed on
  canonical node identity) makes each unique sub-problem computed exactly
  once.

\item \textbf{Incremental subtraction.}  The new \texttt{ComputeDeadlocked}
  procedure starts with the full reachable set and subtracts states that
  have a successor under each event, terminating early when the set
  becomes empty.  Since most states have successors, the MDD shrinks
  rapidly after the first few events, making subsequent events trivially
  cheap.
\end{enumerate}

Together, these reduce $N = 32$ deadlock detection from $> 10$ minutes
to 21\,ms---a speedup of over $28{,}000\times$.

%----------------------------------------------------------------------
\subsection{Phase~3d: Scaling Study}
%----------------------------------------------------------------------

We generated dining philosopher systems for $N = 2$ to $1{,}024$ and
measured the number of MDD nodes, saturation time, deadlock detection
time, and total wall-clock time (including LTS extraction).

\begin{center}
\begin{tabular}{rrrrrr}
\toprule
\textbf{$N$} & \textbf{Processes} & \textbf{MDD Nodes} & \textbf{Saturation} & \textbf{Deadlock Det.} & \textbf{Total} \\
\midrule
2    & 4     & 9     & 0.2\,ms  & 15\,ms   & 1.6\,s \\
4    & 8     & 18    & 2.7\,ms  & 2.4\,ms  & 1.3\,s \\
8    & 16    & 36    & 0.2\,ms  & 4.7\,ms  & 1.3\,s \\
16   & 32    & 72    & 0.3\,ms  & 10\,ms   & 1.5\,s \\
32   & 64    & 144   & 0.4\,ms  & 21\,ms   & 2.1\,s \\
64   & 128   & 288   & 0.6\,ms  & 73\,ms   & 4.0\,s \\
128  & 256   & 576   & 7.4\,ms  & 172\,ms  & 10.0\,s \\
256  & 512   & 1{,}152  & 1.7\,ms  & 911\,ms  & 31.7\,s \\
512  & 1{,}024  & 2{,}304  & 3.4\,ms  & 3.8\,s   & 114.8\,s \\
1{,}024 & 2{,}048  & 4{,}608  & 7.1\,ms  & 19.3\,s  & 430.6\,s \\
\bottomrule
\end{tabular}
\end{center}

The $N = 1{,}024$ system has 2{,}048 concurrent processes with 2{,}048
synchronization channels.  The reachable state space has
$4^{2048} \approx 10^{1233}$ states, yet the MDD representation uses
only 4{,}608 nodes.  Deadlock is detected in 19 seconds; the total
7.2 minutes is dominated by LTS extraction (parsing and compiling
1{,}024 CSP process instances).

Figure~\ref{fig:scaling} plots the three main costs on a log--log scale.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={$N$ (number of philosophers)},
    ylabel={Time (seconds)},
    xmode=log,
    ymode=log,
    log basis x=2,
    xtick={2,4,8,16,32,64,128,256,512,1024},
    xticklabels={2,4,8,16,32,64,128,256,512,1k},
    ymin=1e-4, ymax=1e3,
    width=0.85\textwidth,
    height=0.55\textwidth,
    grid=both,
    grid style={line width=.1pt, draw=gray!20},
    major grid style={line width=.2pt,draw=gray!50},
    legend pos=north west,
    legend style={font=\small},
]

% Total time
\addplot[black, mark=*, thick] coordinates {
    (2, 1.56) (4, 1.34) (8, 1.26) (16, 1.54) (32, 2.09)
    (64, 4.03) (128, 10.0) (256, 31.7) (512, 114.8) (1024, 430.6)
};
\addlegendentry{Total (incl.\ LTS extraction)}

% Deadlock detection time
\addplot[red, mark=triangle*, thick] coordinates {
    (2, 0.015) (4, 0.0024) (8, 0.0047) (16, 0.010) (32, 0.021)
    (64, 0.073) (128, 0.172) (256, 0.911) (512, 3.78) (1024, 19.3)
};
\addlegendentry{Deadlock detection}

% Saturation time
\addplot[blue, mark=square*, thick] coordinates {
    (2, 0.0002) (4, 0.0027) (8, 0.0002) (16, 0.0003) (32, 0.0004)
    (64, 0.0006) (128, 0.0074) (256, 0.0017) (512, 0.0034) (1024, 0.0071)
};
\addlegendentry{Saturation}

\end{axis}
\end{tikzpicture}
\caption{MDD saturation scaling for dining philosophers ($N = 2$ to
$1{,}024$).  Saturation (blue) is essentially free ($< 10$\,ms).
Deadlock detection (red) grows polynomially ($\sim N^{2.4}$).
Total time (black) is dominated by LTS extraction at large $N$.}
\label{fig:scaling}
\end{figure}

\subsubsection{Analysis}

Key observations:
\begin{itemize}
\item \textbf{MDD nodes grow linearly}: $\text{nodes} = 4.5N$, reflecting
  the regular ring structure.  The greedy BFS level ordering keeps all
  synchronization event spans at 1--2 levels.
\item \textbf{Saturation is essentially free}: Under 10\,ms for all
  $N$ tested, even though the state space grows as $4^{2N}$.
\item \textbf{Deadlock detection grows as $O(N^{2.4})$}: The memoized
  incremental subtraction scales polynomially rather than exponentially.
  At $N = 1{,}024$, it takes 19\,s---well within practical limits.
\item \textbf{LTS extraction is the new bottleneck}: Parsing and
  compiling each CSP process instance takes $\sim 0.4$\,s.  For
  $N = 1{,}024$ (2{,}048 instances of 2 process types), this dominates.
  Caching compiled process templates would eliminate this cost.
\end{itemize}

%======================================================================
\section{References}
%======================================================================

\begin{thebibliography}{99}

\bibitem{hoare1985}
C.~A.~R. Hoare,
\emph{Communicating Sequential Processes},
Prentice Hall, 1985.
Available online: \url{http://www.usingcsp.com/}

\bibitem{roscoe2010}
A.~W. Roscoe,
\emph{Understanding Concurrent Systems},
Springer, 2010.

\bibitem{fdr4}
T.~Gibson-Robinson, P.~Armstrong, A.~Boulgakov, A.~W. Roscoe,
``FDR3: A Modern Refinement Checker for CSP,''
\emph{Proc.\ TACAS}, LNCS 8413, pp.~187--201, 2014.
FDR4 documentation: \url{https://cocotec.io/fdr/}

\bibitem{martin1990}
A.~J. Martin,
``Programming in VLSI: From Communicating Processes to Delay-Insensitive Circuits,''
\emph{Developments in Concurrency and Communication}, UT Year of Programming Series,
Addison-Wesley, pp.~1--64, 1990.

\bibitem{martin2006}
A.~J. Martin, M.~Nystr\"om,
``Asynchronous Techniques for System-on-Chip Design,''
\emph{Proc.\ IEEE}, vol.~94, no.~6, pp.~1089--1120, June 2006.

\bibitem{josephs1995}
M.~B. Josephs, J.~T. Udding,
``An Overview of DI Algebra,''
\emph{Proc.\ Hawaii International Conference on System Sciences}, 1995.

\bibitem{por}
P.~Godefroid,
\emph{Partial-Order Methods for the Verification of Concurrent Systems},
LNCS 1032, Springer, 1996.

\bibitem{mcmillan1993}
K.~L. McMillan,
\emph{Symbolic Model Checking},
Kluwer Academic Publishers, 1993.

\bibitem{supercomb}
A.~W. Roscoe,
``Compiling Shared-Nothing CSP to Supercombinators: A Progress Report,''
Technical report, University of Oxford, 2013.

\bibitem{ciardo2001}
G.~Ciardo, G.~L\"uttgen, R.~Siminiceanu,
``Saturation: An Efficient Iteration Strategy for Symbolic State-Space Generation,''
\emph{Proc.\ TACAS}, LNCS 2031, pp.~328--342, 2001.

\end{thebibliography}

\end{document}
